Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2023-06-05 Manuscript/Manuscript.Rmd
Ms: whole-text edits, modelling section, discussion end

Nitric oxide (NO) generated by nitric-oxide synthase (NOS) is a key regulator of animal physiology.
Here we uncover a role for NO in gating a light-avoidance circuit and mediating short-term memory of UV exposure.
We studied UV/violet light avoidance mediated by brain ciliary photoreceptors (cPRCs) in larval *Platynereis dumerilii*, a marine annelid.
We found NOS expressed in interneurons (INNOS) postsynaptic to cPRCs.
UV stimulation of cPRCs leads to INNOS activation and NO production.
NO signals retrogradely to trigger the sustained post-stimulus activation of cPRCs through an unconventional guanylyl cyclase.
This late activation inhibits serotonergic ciliomotor neurons via projection interneurons.
In *NOS* mutants, retrograde signalling, circuit output and UV-avoidance behaviour are defective.
By mathematical modelling, we recapitulate phototransduction and circuit dynamics in both wild-type and mutant larvae.
Our results reveal how NO-mediated retrograde signalling gates a synaptic circuit and induces short-term memory of UV exposure to orchestrate light-avoidance behaviour.

@@

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized by nitric oxide synthase (NOS) from L-arginine is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the Ca2+/calmodulin-dependent activation of NOS following calcium increase, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC).
This can occur cell-autonomously or in other cells leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is calcium dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Aso2019; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018; @Aso2019] .

@@

To investigate NO function in neural circuit dynamics and behaviour, here we study larvae of the marine annelid model *Platynereis dumerilii* [@ozpolat2021].
*Platynereis* has emerged as a model for systems neuroscience where behaviour and activity imaging can be combined with genetic manipulations.
A whole-body synaptic connectome and gene expression atlases are also available [@Vergara2021; @Veraszto2017; @veraszto2020; @ozpolat2021].

Here we uncovered an essential function for NO signalling in larval UV/violet-light avoidance behaviour.
In *Platynereis* larvae, UV/violet avoidance is mediated by brain ciliary photoreceptor cells (cPRCs) and is characterised by downward swimming [@Veraszto2018].
The cPRCs express a ciliary-type opsin, c-opsin1 [@Arendt2004] that forms a UV-absorbing bistable photopigment with an absorption maximum around 384 nm [@tsukamoto2017, @Veraszto2018, @Veedin2021].
Upon UV/violet exposure, the cPRC show a characteristic biphasic calcium response that is *c-opsin1* dependent.
UV/violet avoidance is also *c-opsin1*-dependent and is defective in *c-opsin1* mutants [@Veraszto2018].
Here we show that *Platynereis* *NOS* is expressed in interneurons of the cPRC circuit and is required for UV/violet-avoidance.
By combining calcium imaging across the fully-mapped cPRC circuit [@Veraszto2018] with genetic perturbations and mathematical modelling, we describe how NO tunes circuit dynamics through non-synaptic retrograde signalling to cPRCs.
This delayed neuroendocrine feedback integrates UV/violet exposure to induce a short-term memory manifested in altered circuit activity and an aversive behavioural response.

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences (Figure 1---figure supplement 1).
To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis.
In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region (Figure 1D and Figure 1---figure supplement 2).
*NOS* was also expressed in the region of the visual eyes (adult eyes) and the pigmented eyespots (Figure 1---figure supplement 2).
The four apical organ cells, but not the eyes, were also labelled with a *NOS* reporter construct driving palmitoylated-tdTomato (Figure 1E).
This reporter also revealed the axonal projections of these central NOS-expressing neurons.
The position and morphology of the four *NOS+* cells allowed us to identify the same four cells as four interneurons (INNOS) in our three-day-old whole-body *Platynereis* volume EM data [@williams2017; @veraszto2020] (Figure 1B,C).
In the synaptic connectome, the INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Figure 1C,F).
The projections of INNOS cells are segregated into input and output compartments, with cPRCs inputs occurring in the dense neurosecretory plexus.
INNOS output synapses form in the more ventral projection region (Figure 1---figure supplement 3).
INRGW neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the prototroch ciliary band and the cholinergic MC ciliomotor neurons (Figure 1C,F) [@Veraszto2017].

```{r echo=FALSE,fig.cap="**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** Scanning electron microscopy image of a three-day-old *Platynereis* larva. **(B, C)** Volume rendering of the neuron types (cPRC, INNOS, INRGW, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. **(D)** Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(E)** Expression of a *NOS* reporter (NOSp::palmi-3xHA: magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(F)** Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses."}

@@

```{r echo=FALSE,fig.cap="**Figure 2. NO produced by UV/violet stimulation to cPRCs.** **(A)** DAF-FM fluorescence in the region of the neurosecretory neuropil. The white line indicates the outline of the larva. The dashed line corresponds to the area where fluorescence was quantified. Circles indicate the location of cPRC and control stimulation. cPRCs are marked by thin lines. **(B)** Changes in DAF-FM fluorescence before and during 405 nm light stimulation. **(C)** Changes in DAF-FM fluorescence over time during 405 nm stimulation of the cPRCs (green) or a control area (ctr: gray). The purple box indicate the duration of 405 nm stimulation. Individual traces normalized (ΔF/F0) are shown as thin lines. Thick lines show the mean value with 0.95 confidence intervals. N = 9 larvae for control and 11 for cPRC stimulation."}

@@

We next tested whether NO signalling is required for UV/violet avoidance.
To achieve this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method.
We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to frame-shift and an early stop codon and thus likely representing null alleles (Figure 3---figure supplement 1A).
We could establish a homozygous line for both mutations indicating that NOS is not an essential gene in *Platynereis*.
To quantify UV avoidance, we recorded the trajectories of freely swimming wild type and mutant larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm UV light (Figure 3A and Figure 3---figure supplement 1B).
As previously shown, wild-type larvae swim downward following non-directional UV/violet light stimulation [@Veraszto2018].
In contrast, both two- and three-day-old homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 3A, B and Figure 3---figure supplement 1B,C).
This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour [@Veraszto2018].
Wild type but not mutant larvae also showed an increase in swimming speed under UV light that may be due to downward swimming trajectories (swimming in the direction of gravity) (Figure 3B and Figure 3---figure supplement 1C).
We also tested directional phototaxis, by exposing larvae to 480 nm directional collimated light from the top of the column.
Three-day-old but not two-day-old *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function for *NOS* in the visual eyes that mediate three-day-old phototaxis [@Randel2014a] (Figure 3D and Figure 3---figure supplement 1G).
To distinguish between an acute and developmental function of NOS in light responses, we next tested larvae exposed to the NOS inhibitor L-NAME.
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a dose-dependent inhibition of UV avoidance.
In contrast, phototaxis was not affected (Figure 3C, E).
Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system, reminiscent of the function of NO signalling in *Drosophila* eye development [@Gibbs1998a].

```{r echo=FALSE,fig.cap="**Figure 3. NOS is required for UV avoidance in *Platynereis* larvae.**  **(B)** Swimming trajectories of wild type (WT, n=32) and NOS mutant (NOSΔ11/Δ11, n=26 and NOSΔ23/Δ23, n=47) three-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec after the onset of 395 nm stimulation from the side. **(B)** Vertical position of batches of wild type and mutant three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. **(C)** Vertical position of batches of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. **(D)** Vertical displacement in 30 sec bins of wild type and mutant (NOSΔ11 and NOSΔ23) three-day-old larvae stimulated with 395 nm light from side, 488 nm light from the top and 395 nm light from the top. **(E)** Vertical displacement in 30 sec bins of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top."}

@@

To investigate how NO signalling alters the dynamics of the cPRC circuit, we carried out calcium imaging experiments.
We ubiquitously expressed the calcium sensor GCaMP6s in larvae and imaged calcium signals during 405 nm stimulation of the cPRCs.
As we have shown previously, a 20-sec local stimulation of cPRC cilia lead to a transient increase in cPRC calcium levels, followed by a transient decrease [@Veraszto2018].
After \~20-sec, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus -- a response that may involve depolarisation (Figure 4A, B).
This activation phase occurs after the 20 sec stimulation period and is likely due to a delayed neuroendocrine feedback [@Veraszto2018].
To determine whether NO mediates such a feedback, we repeated the experiment in *NOS*-mutant larvae.
While we detected the initial activation phase followed by inhibition, in homozygous *NOS*-mutants for both CRISPR alleles this was not followed by delayed activation.
Instead, calcium levels dropped to a low steady-state level (Figure 4A, B).
We thus identified a requirement for NO signalling in the late-phase activation of cPRCs.

@@

We aimed next to identify the NO-receptor in the cPRCs.
NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PFAM domain: PF00211).
NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production.
Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit [@Veraszto2017].
Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020].
To identify NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs (Figure 4---figure supplement 1 and 2).
To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup.
In agreement with Moroz et al. [@Moroz2020], we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans.
The 15 *Platynereis* sequences belonged to several deeply diverged clades in the phylogenetic tree (Figure 4---figure supplement 1 and 2).
To characterise the expression of NIT-GCs, we used previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017].
Among the 15 NIT-GCs, two showed high and specific expression in the cPRCs and one was expressed in the INNOS cells (Figure 4---figure supplement 2).
In the single-cell data, we could identify the cPRCs by the specific expression of *c-opsin1* and the *pedal-peptide2 neuropeptide precursor* (*MLD proneuropeptide*), previously described cPRC markers [@Arendt2004; @williams2017] (Figure 4---figure supplement 3A).
The INNOS cells were identified by NOS expression and spatial mapping in the brain [@Achim2015].
We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequence, *NIT-GC1* and *NIT-GC2*.
To confirm the single-cell data, we first carried out in situ hybridisation chain reaction (HCR) with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes were specifically expression in the four cPRCs, as confirmed by co-labeling with an acetylated α-tubulin antibody and with an HCR probe against *pedal peptide 2/MLD proneuropeptide* (Figure 4C, D and Figure 4---figure supplement 3A-C).
To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins.
In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus (Figure 4E).
Co-immunostaining with the rabbit NIT-GC1 antibody and a custom rat antibody raised against *Platynereis* NOS revealed the localisation of both proteins in dots in the neurosecretory neuropil (Figure 4---figure supplement 3D).
In contrast, NIT-GC2 specifically labelled the ramified sensory cilia of the cPRCs (Figure 4F).
These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

To further characterise these atypical guanylyl cyclases, we focused on NIT-GC1 and carried out in vitro experiments.
In bacteria, NIT domains are thought to regulate cellular functions in response to intra- or extracellular nitrate and nitrite.
NIT-GC1 has a NIT domain and a highly conserved cyclase domain that is expected to catalyse cGMP synthesis (Figure 4G).
The NIT domain may render NIT-GC1 dependent on NO signals.
To test this, we co-expressed the cGMP indicator Green cGull [@Matsuda2017] and NIT-GC1 in cultured COS-7 (monkey kidney) cells, a cell line with minimal endogenous sGC activity.
For balanced expression, we used a single plasmid with the two open-reading frames separated by the 2A self-cleaving peptide (Figure 4H).
Application of the NO donor SNAP lead to increased cGMP levels, an effect we did not observe when cells were exposed to DMSO or when Green cGull was expressed alone (Figure 4I-K).
To test whether cGMP production is dependent on the NIT domain, we also tested a deletion construct of NIT-GC1 lacking the NIT domain (Figure 4G).
Cells expressing this construct and Green cGull did not show increased cGMP levels when exposed to SNAP (Figure 4L).
These results indicate that NIT-GC1 is able to catalyse cGMP production in an NO-dependent manner and this function requires the NIT domain.
These results establish NIT-GC1 as a biochemical NO sensor.

@@

```{r echo=FALSE,fig.cap="**Figure 4. NOS and two NIT-GCs shape calcium signals during cPRC UV/violet response.** **(A, B)** GCaMP6s signals in cPRCs in wild type and *NOS* mutant (A, NOSΔ11/Δ11, B, NOSΔ23/Δ23) larvae during 405 nm light stimulation. **(C, D)** In situ HCR for (C) *NIT-GC1* and (D) *NIT-GC2* (magenta) in three-day-old *Platynereis* larvae. Larvae were co-stained with an antibody agains acetylated α-tubulin to label cPRC cilia and the neuropil (green). **(E, F)** Immunostaining for (E) NIT-GC1 and (F) NIT-GC2 (magenta), co-stained for acetylated α-tubulin (green). **(G)** The domain structure of *Platynereis* NIT-GC1 and the truncated NIT-GC1ΔNIT protein lacking the NIT domain. A predicted transmembrane region (TM) is shown in grey. **(H)** Schematic of the cell-based assay to detect cGMP production following the addition of an NO donor SNAP or DMSO as control. **(I-L)** Green cGull fluorescence over time for the four conditions tested. Individual responses and their mean with 0.95 confidence interval are shown (n > 6 cells). Intensities are normalized (ΔF/F0). The indicated chemicals were added at 2 min after the start of imaging (grey bars). **(M, N)** GCaMP6s signals in cPRCs in (M) NIT-GC1 and (N) NIT-GC2 morphant larvae during 405 nm light stimulation. Individual responses and their mean with 0.95 confidence interval are shown."}

>>>>

Nitric oxide (NO) produced by nitric-oxide synthase (NOS) is a key regulator of animal physiology.
Here we uncover a function for NO in the integration of UV exposure and the gating of a UV-avoidance circuit.
We studied UV/violet avoidance mediated by brain ciliary photoreceptors (cPRCs) in larvae of the annelid *Platynereis dumerilii*. In the larva, NOS is expressed in interneurons (INNOS) postsynaptic to cPRCs.
UV stimulation of cPRCs triggers INNOS activation and NO production.
NO signals retrogradely to cPRCs to induce their sustained post-stimulus activation through an unconventional guanylyl cyclase.
This late activation inhibits serotonergic ciliomotor neurons to induce downward swimming.
In *NOS* mutants, retrograde signalling, circuit output and UV avoidance are defective.
By mathematical modelling, we recapitulate phototransduction and circuit dynamics in wild-type and mutant larvae.
Our results reveal how NO-mediated retrograde signalling gates a synaptic circuit and induces short-term memory of UV exposure to orchestrate light-avoidance behaviour.

@@

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized from L-arginine by nitric oxide synthase (NOS) is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the Ca2+/calmodulin-dependent activation of NOS, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC).
This can occur cell-autonomously or in other cells leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is calcium dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Aso2019; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018; @Aso2019] .

@@

To investigate NO function in neural circuit dynamics and behaviour, here we study larvae of the marine annelid model *Platynereis dumerilii* [@ozpolat2021].
*Platynereis* has emerged as a model for systems neuroscience with a toolkit that enables the combination of behavioural analysis and neuronal activity imaging with genetic manipulations.
A whole-body synaptic connectome and gene expression atlases are also available and can be integrated with functional approaches in live animals thanks to the cellular-level stereotypy of larvae of the same developmental stage [@Vergara2021; @veraszto2017; @veraszto2020; @ozpolat2021].

Here we uncovered an essential function for NO signalling in larval UV/violet-light avoidance behaviour.
In *Platynereis* larvae, UV/violet avoidance is mediated by brain ciliary photoreceptor cells (cPRCs) and is characterised by downward swimming [@Veraszto2018].
The cPRCs express a ciliary-type opsin, c-opsin1 [@Arendt2004] that forms a UV-absorbing bistable photopigment with an absorption maximum around 384 nm [@tsukamoto2017; @Veraszto2018; @Veedin2021].
Upon UV/violet exposure, the cPRC show a characteristic biphasic calcium response that is *c-opsin1* dependent.
UV/violet avoidance is also *c-opsin1*-dependent and is defective in *c-opsin1* mutants [@Veraszto2018].
Here we show that *Platynereis* *NOS* is expressed in interneurons of the cPRC circuit and is required for UV/violet-avoidance.
By combining calcium imaging across the fully-mapped cPRC circuit [@Veraszto2018] with genetic perturbations and mathematical modelling, we describe how NO tunes circuit dynamics through non-synaptic retrograde signalling to cPRCs.
This delayed neuroendocrine feedback integrates UV/violet exposure to induce a short-term memory manifested in altered circuit activity and an aversive behavioural response.

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of bilaterian NOS sequences (Figure 1---figure supplement 1).
To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis.
In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region (Figure 1D and Figure 1---figure supplement 2).
*NOS* was also expressed in the region of the visual eyes (adult eyes) and the pigmented eyespots (Figure 1---figure supplement 2).
The four apical organ cells, but not the eyes, were also labelled with a *NOS* reporter construct driving palmitoylated-tdTomato (Figure 1E).
This reporter also revealed the axonal projections of these central NOS-expressing neurons.
The position and morphology of the four *NOS+* cells allowed us to identify the same four cells as four interneurons (INNOS) in our three-day-old whole-body *Platynereis* volume EM data [@williams2017; @veraszto2020] (Figure 1B,C).
In the synaptic connectome, the INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGWa interneurons, which are also cPRC targets (Figure 1C,F).
The projections of INNOS cells are segregated into input (dendritic) and output (axonal) compartments, with cPRC inputs occurring on the dendritic part in the dense neurosecretory plexus.
INNOS output synapses form in the more ventral projection region (Figure 1---figure supplement 3).
INRGWa neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the prototroch ciliary band and the cholinergic MC ciliomotor neurons (Figure 1C,F) [@veraszto2017].

```{r echo=FALSE,fig.cap="**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** Scanning electron microscopy image of a three-day-old *Platynereis* larva. **(B, C)** Volume rendering of the neuron types (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. **(D)** Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(E)** Expression of a *NOS* reporter (NOSp::palmi-3xHA: magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(F)** Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses."}

@@

```{r echo=FALSE,fig.cap="**Figure 2. NO produced by UV/violet stimulation to cPRCs.** **(A)** DAF-FM fluorescence in the region of the neurosecretory neuropil. The white line indicates the outline of the larva. The dashed line corresponds to the area where fluorescence was quantified. Circles indicate the location of cPRC and control stimulation. cPRCs are marked by thin lines. **(B)** Changes in DAF-FM fluorescence before and during 405 nm light stimulation. **(C)** Changes in DAF-FM fluorescence over time during 405 nm stimulation of the cPRCs (green) or a control area (ctr: gray). The purple box indicates the duration of 405 nm stimulation. Individual traces normalized (ΔF/F0) are shown as thin lines. Thick lines show the mean value with 0.95 confidence intervals. N = 9 larvae for control and 11 for cPRC stimulation."}

@@

We next tested whether NO signalling regulates UV/violet avoidance.
To achieve this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method.
We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both frame-shift mutations leading to an early stop codon and thus likely representing null alleles (Figure 3---figure supplement 1A).
We could establish a homozygous line for both mutations indicating that NOS is not an essential gene in *Platynereis*.
To quantify UV avoidance, we recorded the trajectories of freely swimming wild type and mutant larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm UV light (Figure 3A and Figure 3---figure supplement 1B).
As previously shown, wild-type larvae swim downward following non-directional UV/violet light stimulation [@Veraszto2018].
In contrast, both two- and three-day-old homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 3A, B and Figure 3---figure supplement 1B,C).
This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae [@Veraszto2018] and reveals a requirement for NOS in UV-avoidance behaviour.
Wild type but not mutant larvae also showed an increase in swimming speed under UV light that may be due to downward swimming trajectories (swimming in the direction of gravity) (Figure 3B and Figure 3---figure supplement 1C).
We also tested directional phototaxis, by exposing larvae to 480 nm directional collimated light from the top of the column.
Three-day-old but not two-day-old *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function for *NOS* in the visual eyes that mediate three-day-old phototaxis [@Randel2014a] (Figure 3D and Figure 3---figure supplement 1G).

To distinguish between an acute and developmental function of NOS in light responses, we next tested larvae exposed to the NOS inhibitor L-NAME.
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a dose-dependent inhibition of UV avoidance.
In contrast, phototaxis was not affected (Figure 3C, E).
Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system, reminiscent of the function of NO signalling in *Drosophila* eye development [@Gibbs1998a].

```{r echo=FALSE,fig.cap="**Figure 3. NOS is required for UV avoidance in *Platynereis* larvae.**  **(B)** Swimming trajectories of wild type (WT, n=32) and NOS mutant (NOSΔ11/Δ11, n=26 and NOSΔ23/Δ23, n=47) three-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec after the onset of 395 nm stimulation from the side. **(B)** Vertical position of batches of wild type and mutant three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. **(C)** Vertical position of batches of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. **(D)** Vertical displacement in 30 sec bins of wild type and mutant (NOSΔ11 and NOSΔ23) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. **(E)** Vertical displacement in 30 sec bins of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top."}

@@

To investigate how NO signalling alters the dynamics of the cPRC circuit, we carried out calcium imaging experiments.
We ubiquitously expressed the calcium sensor GCaMP6s in larvae and imaged calcium signals during 405 nm stimulation of the cPRCs.
As we have shown previously, a 20-sec local stimulation of cPRC cilia led to a transient increase in cPRC calcium levels, followed by a transient decrease [@Veraszto2018].
After \~20-sec, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus -- a response that may involve depolarisation (Figure 4A, B).
This activation phase occurs after the 20 sec stimulation period and is likely due to a delayed neuroendocrine feedback [@Veraszto2018].
To determine whether NO mediates such a feedback, we repeated the experiment in *NOS*-mutant larvae.
While we detected the initial activation phase followed by inhibition, in homozygous *NOS*-mutants for both CRISPR alleles this was not followed by delayed activation.
Instead, calcium levels dropped to a low steady-state level (Figure 4A, B).
We thus identified a requirement for NO signalling in the late-phase activation of cPRCs.

@@

We aimed next to identify the NO receptor in the cPRCs.
NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PFAM domain: PF00211).
NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production.
Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit [@veraszto2017].
Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020].
To identify NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs (Figure 4---figure supplement 1 and 2).
To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup.
In agreement with Moroz et al. [@Moroz2020], we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans.
The 15 *Platynereis* sequences belonged to several deeply diverged clades in the phylogenetic tree (Figure 4---figure supplement 1 and 2).

To characterise the expression of NIT-GCs, we used previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017].
Among the 15 NIT-GCs, two showed high and specific expression in the cPRCs and one was expressed in the INNOS cells (Figure 4---figure supplement 2).
In the single-cell data, we could identify the cPRCs by the specific expression of *c-opsin1* and the *pedal-peptide2 neuropeptide precursor* (*MLD proneuropeptide*), previously described cPRC markers [@Arendt2004; @williams2017] (Figure 4---figure supplement 3A).
The INNOS cells were identified by NOS expression and spatial mapping in the brain [@Achim2015].
We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequence, *NIT-GC1* and *NIT-GC2*.
To confirm the single-cell data, we first carried out in situ hybridisation chain reaction (HCR) with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes were specifically expression in the four cPRCs, as confirmed by co-labeling with an acetylated α-tubulin antibody and with an HCR probe against *pedal peptide 2/MLD proneuropeptide* (Figure 4C, D and Figure 4---figure supplement 3A-C).
To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins.
In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior neurosecretory plexus (Figure 4E).
Co-immunostaining with the rabbit NIT-GC1 antibody and a custom rat antibody raised against *Platynereis* NOS revealed the localisation of both proteins in dots in the neurosecretory neuropil (Figure 4---figure supplement 3D).
In contrast, NIT-GC2 specifically labelled the ramified sensory cilia of the cPRCs (Figure 4F).
These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

To further characterise these atypical guanylyl cyclases, we focused on NIT-GC1 and carried out in vitro experiments.
In bacteria, NIT domains are thought to regulate cellular functions in response to intra- or extracellular nitrate and nitrite.
NIT-GC1 has a NIT domain and a highly conserved cyclase domain that is expected to catalyse cGMP synthesis (Figure 4G).
The NIT domain may render NIT-GC1 dependent on NO signals.
To test this, we co-expressed the cGMP indicator Green cGull [@Matsuda2017] and NIT-GC1 in cultured COS-7 (monkey kidney) cells, a cell line with minimal endogenous sGC activity [@Matsuda2017].
For balanced expression, we used a single plasmid with the two open-reading frames separated by the 2A self-cleaving peptide (Figure 4H).
Application of the NO donor SNAP lead to increased Green cGull fluorescence, an effect we did not observe when cells were exposed to DMSO or when Green cGull was expressed alone (Figure 4I-K).
To test whether this effect is dependent on the NIT domain, we also tested a deletion construct of NIT-GC1 lacking the NIT domain (Figure 4G).
Cells expressing this construct and Green cGull did not show an increased fluorescence of the cGMP reporter when exposed to SNAP (Figure 4L).
These results indicate that NIT-GC1 is able to catalyse cGMP production in an NO-dependent manner and this function requires the NIT domain.
These results establish NIT-GC1 as a biochemical sensor of NO or its derivatives.

@@

```{r echo=FALSE,fig.cap="**Figure 4. NOS and two NIT-GCs shape calcium signals during cPRC UV/violet response.** **(A, B)** GCaMP6s signals in cPRCs in wild type and *NOS* mutant (A, NOSΔ11/Δ11, B, NOSΔ23/Δ23) larvae during 405 nm light stimulation. **(C, D)** In situ HCR for (C) *NIT-GC1* and (D) *NIT-GC2* (magenta) in three-day-old *Platynereis* larvae. Larvae were co-stained with an antibody against acetylated α-tubulin to label cPRC cilia and the neuropil (green). **(E, F)** Immunostaining for (E) NIT-GC1 and (F) NIT-GC2 (magenta), co-stained for acetylated α-tubulin (green). **(G)** The domain structure of *Platynereis* NIT-GC1 and the truncated NIT-GC1ΔNIT protein lacking the NIT domain. A predicted transmembrane region (TM) is shown in grey. **(H)** Schematic of the cell-based assay to detect cGMP production following the addition of an NO donor SNAP or DMSO as control. **(I-L)** Green cGull fluorescence over time for the four conditions tested. Individual responses and their mean with 0.95 confidence interval are shown (n > 6 cells). Intensities are normalized (ΔF/F0). The indicated chemicals were added at 2 min after the start of imaging (grey bars). **(M, N)** GCaMP6s signals in cPRCs in (M) NIT-GC1 and (N) NIT-GC2 morphant larvae during 405 nm light stimulation. Individual responses and their mean with 0.95 confidence interval are shown."}
